---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r setoptions, echo=TRUE}
library("knitr")
require('knitr')
opts_chunk$set(echo=TRUE,results=TRUE)
indata<-read.csv("activity.csv")
# check summary of the data
head(indata)
summary(indata)
colnames(indata)
str(indata)

```
## Calculate mean total number of steps taken per day
```{r meantotal,echo=TRUE}
indata<-subset(indata,!(is.na(indata$steps)))
                                 
y<-as.Date(indata$date,"%Y-%m-%d")
# average per day
aggtotal<-aggregate(indata$steps,by=list(y),FUN=sum)
hist(as.vector(aggtotal[,2]),breaks=10,main="Histogram of total steps per day",xlab="Total steps per day", ylab="Frequency")
#Mean and median steps per day

aggmeanday<-aggregate(indata$steps,by=list(y),FUN=mean)
aggmedianday<-aggregate(indata$steps,by=list(y),FUN=median)
names(aggmeanday)<-c("Date","Mean_steps")
names(aggmedianday)<-c("Date","Median_steps")
aggday<-merge(aggmeanday,aggmedianday,by="Date")

# mean steps per interval
agginterval<-aggregate(indata$steps,by=list(indata$interval),FUN=mean)


```

## What is mean total number of steps taken per day?
Mean and Median total number of steps per day is listed below;
```{r tablevalues, echo=TRUE}
print(aggday,row.names=FALSE)
```

## What is the average daily activity pattern?

```{r pattern, echo=TRUE}
# average daily activity vs interval
agginterval<-aggregate(indata$steps,by=list(indata$interval),FUN=mean)
plot(agginterval[,1],agginterval[,2],type="l",main="Average daily activity pattern",xlab="five minute interval",ylab="Average steps")
index<-which.max(agginterval[,2])

```
The average daily activity pattern is maximum during the interval `r format(agginterval[index,1])`.

## Imputing missing values
```{r impute,echo=TRUE}
indatafull<-read.csv("activity.csv")
a<-is.na(indatafull$steps)
nimpute<-sum(a)
#fill missing values by means
indatafull$steps[is.na(indatafull$steps)]<-mean(indatafull$steps,na.rm=TRUE)
# average per day
yall<-as.Date(indatafull$date,"%Y-%m-%d")
aggtotalall<-aggregate(indatafull$steps,by=list(yall),FUN=sum)
hist(as.vector(aggtotalall[,2]),breaks=10,main="Histogram of total steps per day after Imputing",xlab="Total steps per day", ylab="Frequency")
str(aggtotalall)
aggallmeanday<-aggregate(indatafull$steps,by=list(yall),FUN=mean)
aggallmedianday<-aggregate(indatafull$steps,by=list(yall),FUN=median)
names(aggallmeanday)<-c("Date","Mean steps per day")
names(aggallmedianday)<-c("Date","Median steps per day")
aggallday<-merge(aggallmeanday,aggallmedianday,by="Date")
```
The number of rows with data missing on steps are `r nimpute`.
After imputing, the mean and the median of the total number of steps per day is  shown below;
```{r imputeday,echo="TRUE"}
print(aggallday,rownames="FALSE")
```


Imputing the missing values, the median steps per day was non-zero for some days. Before imputing, the median steps per day was zero for all days. The mean values did not vary significantly after imputing. 

## Are there differences in activity patterns between weekdays and weekends?
```{r days,echo=TRUE}
library(ggplot2)
# get weekday information
indatafull$day<-weekdays(as.Date(indatafull$date,"%Y-%m-%d"))
#indatafull$period<-ifelse(indatafull$day=="Saturday" | indatafull$day=="Sunday","Weekend","Weekday")
# create factor variable using Weekday information
indatafull$period<-factor(indatafull$day=="Saturday" | indatafull$day=="Sunday")
levels(indatafull$period)[levels(indatafull$period)=="TRUE"]<-"Weekend"
levels(indatafull$period)[levels(indatafull$period)=="FALSE"]<-"Weekday"
# check the new structure of the imputed data after adding new column
str(indatafull)
# compute average activity per interval
aggperiod<-aggregate(indatafull$steps,by=list(indatafull$interval,indatafull$period),FUN=mean)
str(aggperiod)
#plot
qplot(Group.1,x,data=aggperiod,facets=.~Group.2,geom=c("line","smooth"),method="lm",main="Comparison of activity over Weekend and Weekdays",xlab="Interval",ylab="Average steps")


```

Looking at the regression fit lines, the average activity over the weekend appears to be higher than that over the weekdays. The time-series plot during the weekdays shows that the average activity during the early intervals is higher than the later intervals. The time-series plot during the weekends, suggest that the average activity remains similar throughout the day.